Governing through patronage: the bargain for education in decentralized Brazil

### 2020-10-05 paper/paper.tex
push paper

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}. The World Bank refers to as ``learning poverty'': poor learning when access to a schooling already exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} As scholarly focus shifts from public service provision \citep{kaufman_globalization_2001, } to the quality

Raises two related questions. The first one concerns the relationship between the quality of public services and the structure of the bureaucracy \citep{pepinsky_bureaucracy_2017, rasul2018management}.

@@

\subsection*{Presidential coalitionism and patronage}

For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}. Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. \citet{raile_executive_2011, power_optimism_2010}.

In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

@@

@@

Brazil's local level government is composed by over 5 thousand municipalities, each with their own mayor hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}). Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.\footnote{The last municipal elections were in 2016. These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over personnel apppointment. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.

Interviews with secretaries of education and mayors provide evidence that city councilors play an important role in staffing decisions. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. In the words of a set of school administrators in the municipality of I:

    \emph{Educational counselor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J. August 2019.}
\end{quote}

These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff. To verify these claims in a broader set of municipalities, as well as outlining the research design, I apply a theory of legislative vote-buying that illustrates the  employ extensive administrative and electoral data collected by the Brazilian federal government. In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
    \caption{Uneven quality of municipal education. Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores. Averages are weighted by student participation rate.}
    \label{fig:saeb_map}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
    \label{fig:staff_turnover}
    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
\end{figure}

>>>>

% what do I need to bring the audience onboard?
% emphasize the existence of discretion, turnover, political manipulation of public sresources.
% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}. The World Bank refers to as ``learning poverty'': poor learning when access to a schooling already exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} As scholarly focus shifts from public service provision \citep{kaufman_globalization_2001, alesina_public_1999} to the quality of services themselves \citep{}

Raises two related questions. The first one concerns the relationship between the quality of public services and the structure of the bureaucracy \citep{pepinsky_bureaucracy_2017, rasul2018management}. Moving up the causal ladder, the question concerns the relationship between politicians and bureaucratic institutions and how theses personnel decisions \citep{finan_personnel_2015} have downstream consequences for the quality of these services \citep{pepinsky_bureaucracy_2017}

@@

% \subsection*{Presidential coalitionism and patronage}

% For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}. Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. \citet{raile_executive_2011, power_optimism_2010}.

% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

% In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

@@

Brazil's local level of government is composed of over 5 thousand municipalities, each with their own mayorship (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}). Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, open to reelection for both offices.\footnote{The last municipal elections were in 2016. These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over public administration in general, and personnel appointment in particular \citep{akhtari_political_2015,brollo_victor_2017}. As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year, providing a local checks and balances on the executive.

At the federal level, this instituitonal arrangement gives rise to a phenomenon dscribed as presidential coalitionism \citet{raile_executive_2011, power_optimism_2010}. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

Interviews with secretaries of education and mayors provide strong evidence that city councilors play an important role in staffing decisions, despite having no \emph{formal} rights to do so. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. According to accounts by school managers in the municipality of I:

    \emph{Educational councilor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J. August 2019.}
\end{quote}

@@

@@

\begin{figure}[h]
    \centering
    \includegraphics[]{plots/saeb_map.pdf}
    \caption{Uneven quality of municipal education. Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores. Averages are weighted by student participation rate.}
    \label{fig:saeb_map}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\linewidth]{plots/descriptive_staff_turnover.pdf}
    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
    \label{fig:staff_turnover}
\end{figure}
